---
title: "Exploratory Analysis"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Project title

# Gender-based Wage Gap in Professional Sports

### Authors

# Charlie Bond, Sirinya Hiruvinya, Aleah Rosner

### Date

Spring 2024

### Abstract

Our main goal with this project is to explore the difference in salary
between NBA players and WNBA players, which is important because it
points to a larger issue of gender inequality that spans nearly every
career path. To address this question, we will be comparing the salaries
of professional players across both leagues and using our dataset to
analyze trends in salary vs games played, points scored, and other
factors. When doing this, it is important to consider that this data
spans a short time frame (2016-2017), which needs to be taken into
account when making conclusions from our analysis; however, our goal of
bringing light to the issue of the wage gap by applying it to the
popular world of professional sports still stands.

### Keywords

NBA, WNBA, wage gap, gender inequality, salary

### Introduction

A few research questions we hope to explore using this data set include:
How does statistical basketball performance, such as minutes played,
points scored, rebounds, assists, steals, and blocks, differ between the
WNBA and NBA? How do salary-to-performance ratios (salary per point,
salary per rebound, salary per assist, etc.) compare between male and
female basketball players? To what extent does salary vary between WNBA
and NBA players when compared to their performance? Through this
analysis, our intention is to examine how a broad issue of gender
inequality in the workplace is applied to a niche such as elite sports.
This is important because many people don’t acknowledge the fact that
although playing at the most elite level of basketball is a rare feat
for both men and women, the difference in pay between the men’s and
women’s professional leagues is palpable. More broadly, we aim to bring
awareness to the wage gap as a whole by introducing it in relation to a
very popular sport. Acknowledging the wage gap is incredibly important
because only after people are aware of it can changes begin to be made.
By relating the wage gap to sports, it is possible that this issue can
reach a wider audience and more people might be inspired to advocate for
salary transparency in their workspace and careers.

### Related Work

In almost every workplace in the United States, the wage gap impacts how
much women are paid compared to their male counterparts. Although this
number has slowly been increasing since the mid-20th century, women
still only earn 84 cents for every dollar that men earn (Haan, 2024).
Although this degree of this disparity fluctuates between different
career paths, it exists in all but a couple of workplaces. Previous
studies conducted by Forbes advisors have shown that men earn up to 60
percent more than their female counterparts as real estate brokers, and
58 percent more as personal financial advisors (Haan, 2024). These
numbers are even more jarring in the world of professional sports. In
the NBA, male basketball players earn an average of 5 million dollars
annually, which far overshadows female players’ measly annual salaries
of 120,000 dollars (Richard-Craven, 2024). According to Bloomberg
Businessweek, viewership has increased exponentially in the last decade
and a half, with over \$75 million dollars in sponsorships accompanying
this increase in exposure, but players’ salaries have largely stayed the
same (Haque, 2023). This, along with the astonishing wage gap between
the leagues, lends itself to a much larger issue of gender inequality in
the workplace, and is something that needs to be addressed in order to
ensure that female employees are getting justly compensated for their
significant contributions to the workforce - from sports to real estate.

Haan, K. (n.d.). Gender pay gap statistics in 2024 – forbes advisor.
Retrieved May 1, 2024, from
<https://www.forbes.com/advisor/business/gender-pay-gap-statistics/>

Haque, J. (2023, April 26). Women’s basketball is raking in more cash
than ever, but the players aren’t. Bloomberg.Com.
<https://www.bloomberg.com/news/features/2023-04-26/wnba-is-growing-but-players-are>
n-t-getting-a-penny-of-revenue-share

Richard-Craven, M. (n.d.). Addressing the gender pay gap with wnba alum
angel mccoughtry. Forbes. Retrieved May 1, 2024, from
<https://www.forbes.com/sites/mayarichard-craven/2024/03/14/addressing-the-gender-pa>
y-gap-with-wnba-alum-angel-mccoughtry/

### The Dataset

$Where was the data found?$

The data was found by googling “WNBA salary dataset" and "NBA salary
dataset". The data we found for the Women's league is from
herhoopstats.com, and the NBA data was found on gigasheet.com. we also
found data from 2016 by googling "wnba nba salary gap csv".

$Who Collected the data?$

The WNBA data was collected by technologists at Sportradar, a sports
technology company affiliated with HerHoopStats. The NBA data was
uploaded by a community contributor who put the data in CSV format, but
the information is available to the public by the NBA. The 2016 data was
found on github.

$How was the data collected?$

The data is made open to public, and was collected and formatted into a
csv file by technologists from companies such as sportradar.

$Why was the data collected?$

Typically, this data is collected for use by the media (such as sports
announcers) and for sports betting purposes.

$How many observations (rows) are in your data?$

The WNBA has information on 130 players, and the NBA dataset has
information on over 500 players. Combined, there are 630 observations.

$How many features (columns) are in the data?$

The original datasets had 27(WNBA) and 52(NBA) features, but we will
condensed these to the 10 features we felt were most relevant to our
analysis.

What, if any, ethical questions or questions of power do you need to
consider when working with this data?

Because the salary data for WNBA and NBA players is available for the
public to access, there aren’t as many ethical questions to be addressed
in terms of anonymity. However, in order to avoid potential bias in our
analysis and in other people’s exploration of the data, it might be a
good idea to refrain from using specific players’ names in our analysis.
This is especially important as many people already have preconceived
notions about the salaries of the NBA compared to the WNBA, so it is
important to ensure that our analysis and presentation of the data
reflects only the numbers, and is not influenced by previous assumptions
about the relative popularity of the two leagues.

What are possible limitations or problems with this data?

With our ultimate goal being to demonstrate the clear difference in
salaries between men and women, one major drawback of our data is that
it only showcases this disparity in one part of one very niche industry.
Additionally, comparing the data we have on the wage gap in other
industries (such as from the Forbes analysts mentioned above) to the
data we’ve found in professional basketball leagues, the sports figures
we will be analyzing demonstrate an extreme wage gap that is
significantly larger than almost any other industry. For this reason,
it’s important that we acknowledge that the differences we see between
the WNBA and the NBA are more stark than in other areas of work.
Additionally, the data we have on the WNBA has about half the entries
(rows) compared to the NBA. Although we have adequate numbers of data
points for both leagues, this difference could also lead to a slightly
more accurate portrayal of one league compared to the other simply
because more data is available. Keeping in mind both the extremity of
the wage gap in this industry specifically and the difference in data
points is important so that we don’t accidentally end up spreading
misinformation, which could lead to a reversal of the progress that has
been slowly but steadily made to close this gap.

### Implications

By understanding the disparity evident through these statistics, data
analysts and technologists could develop AI or data analysis tools that
could assess player value and their contribution to the games. This
would make it easier to objectively measure a player’s real talent and
provide female players with more fair pay. This helps reduce biases and
provides evident numerical data that could be used to determine a
player's salary.

Designers could take part in creating more compelling campaigns and
narratives to highlight the strengths, achievements, and stories of
female players in the WNBA. Designers play a critical role in shaping
how sports are marketed to the general public, and through strategic and
inclusive branding, they could challenge stereotypes and empower women
in the sports industry. This will garner more attention for women in
sports and WNBA games, which could drive more revenue and empowerment
for the industry.

Our analysis could also highlight the severity of gender inequality to
policymakers; policymakers could implement policies such as the pay
transparency initiative for employers, invest more in women's sports and
representation, and monitor the pay gap in different industries. If
technologists, designers, and policymakers work together, they could
reduce the gender wage gap by coming up with systems and policies that
objectively measure an individual's talents and contributions that
determine their wage.

### Limitations & Challenges

While the datasets are ideal in size and readability, they were
collected from different sources. as we discussed in class, biases and
inaccuracies can skew or alter the accuracy of data, and coming from two
different sources means that they might be affected by different biases.
Additionally, given how niche the professional sports industry is, it is
possible that the trend in the wage gap in this area of work differs
from that of other industries. This poses an additional challenge for
applying our analysis to current trends in the wage gap.

More broadly, one challenge with using this data is that it only
addresses one niche of the gender pay gap, which may be difficult to
generalize to an entire population. These datasets only include the
salaries and pay data of professional basketball players from a very
short time period (2022-2024). In order to address this, we can
acknowledge this specificity in our discussion of our data analysis, or
we could implement other sources (such as news articles) that point out
this pay disparity in more general terms.

### Summary Information

```{r summary info, echo=FALSE, message=FALSE}
source("Scripts/NBA_Summary.R")
```

After performing a preliminary analysis of the data and calculating a
few values, the difference between female and male players' earnings is
astonishing. Beginning with the average salary within each league, we
found that the average salary in the NBA was over
\$`r average_salary_NBA`, while in the WNBA, the average salary was a
mere \$`r average_salary_WNBA`, which is a difference of
\$`r diff_salary_avg`. Taking a closer look, we first compared the
highest and lowest salaries between the leagues. We found that the
maximum NBA salary (48,070,014) was \$`r diff_salary_max` greater than
the highest WNBA salary. Interestingly, the lowest NBA salary (5,849)
was \$ `r diff_salary_min` less than the lowest WNBA salary of 64,154.
We also compared salaries relative to performance metrics such as salary
per point and salary per minute played. Although on average, WNBA
players did tend to score less in games, the money they earned per point
and minute played was still significantly less than their male
counterparts. On average, NBA players made 850,078.40 per point, whereas
WNBA players only earned 18,532.70 for every point they scored (a
difference of 831,546.30). Additionally, NBA players earned 11,637.74
dollars for every minute of play, while female players in the WNBA only
made 6,447.84 dollars (a difference of \$5,189.90).

### Table
```{r aggregatetable, echo = FALSE}
#source("Aggregate_Table.R")
```

This was included to give some general impressions on differences between each league. 
The stats include Average Playtime per game, and Average Salary per Minute, which gives an idea of how the players are compensated when accounting for differences in the structure of each league. 



### Chart 1

Include a chart

<<<<<<< HEAD
##Author: Charlie Bond
```{r chart1, echo = FALSE}
#source("Chart1_performance_stats.R")
```

This chart takes lookes at average salaries for each league and divides them by various perfomance statistics. 
The goal of this is to compare the salaries between the leagues accounting for performance and what is rewarded in each league. 
The ultimate purpose is to show the sheer scale of difference in how players get rewarded in each league, accounting for differences between the leagues. 
=======
Describe why you chose this chart and what information it reveals

Here's an example of how to run an R script inside an RMarkdown file:

\`\`\``{r, echo = FALSE, code = readLines("chart1_example.R")}`
>>>>>>> 8be055ede32408c4ccfe4d7c0c67657f33abc647

### Chart 2

#### Author: Sirinya Hirunviriya

``` {r chart2, echo = FALSE}        
source("CHART2_PALMY.R") 
interactive_plot
```

Purpose: The box plot offers a side-by-side comparison of WNBA and NBA
players, providing insights into the statistical distribution of WNBA
and NBA salaries. By examining statistical data like the interquartile
range (IQR), median, minimum, and maximum salaries, provided in the box
plot, we understand the extent gender pay
gap contributes in professional basketball leagues.

Insights: The box plot underscores that gender pay gap still remains as
a persistent issue in professional basketball. Notably, the maximum
salary of WNBA remains significantly lower than that of the minimum
salary of NBA player, showing a substantial disparity in earning
potential between female and male athlete. This disparity underscores
the significant difference in earning potential between female and male
athletes. Furthermore, the wider Interquartile Range (IQR) among NBA
players also sugguest a broader spectrum of earnings within the league.
This IQR range indicates how much more important market demand, skill,
and reputation are in the NBA than they are in the WNBA, where players
all make comparable salaries despite their skill and reputation.
Additionally, the presence of outliers, such as the NBA player with a
maximum earning of 50 million, further emphasizes the varied earning
profiles within the NBA, while WNBA players tend to earn within a
narrower range. All these statistical data underscores ongoing need for
efforts to reduce the inequality in gender pay gap in sports.

### Chart 3
#### Author: Aleah Rosner

```  {r chart3, echo = FALSE}
source("Chart3_yeartrends.R")
plot(wnba_nba_year_trend_graph)
```

This chart displays the change in average salary in each league between
2016 and 2024. The purpose for showing this is to examine and compare
the trend in calaries among the mens and womens leagues, which can help
us begin to determine whether progress is being made to close the wage
gap. In 2016, the average salary for both is 100% (or 1x) the amount
that it was in 2016, which is why both graphs start at one. The fold
change (amount of increase) was calculated using the formula (2024 avg
salary)/(2016 avg salary). Using this, we determined that the WNBA's
average salary increased to around 1.25 times (125%) the salary from
2016, while the NBA's average salary to 0.94 (or 94%) of its 2016
salary. This data is promising, as it shows that the WNBA's salaries
are, in general, on the rise.
